Constructing a universe for the setoid model

### 2021-01-21 paper.tex
Use \blank for placeholders.

@@

We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
$\Sigma$-types are defined negatively by pairing $\_,\_$ and projections $\pi_1, \pi_2$.
We also have definitional $\eta$-rules for $\Pi, \Sigma$, and $\mathbf{1}$ types.
We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
%
From $W_{\mProp}$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\inU$ for defining $\U$, we also introduce a family
$\inUU$ of binary relations between types in the universe, from which we
then define $\_\sim_{\setoidU}\_$.
{\small\vspace{-0.2em}
\begin{align*}
  & \textsf{data}\ \inU : \mType \to \mType_1 \\
  & \qquad \textsf{bool} : \inU\ \mathbbm{2} \\
  & \qquad \pi : \forall\{A\;A_\sim\;B\;B_\sim\}(a : \inU\;A)
  (b : (x : A) \to \inU\;(B\;x))\\
  & \qquad\ \to \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
  \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
  & \qquad\ \to \inU\;
  (\Sigma\;(f : (x : A) \to B\;x) \\
  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} : A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1))) \\
  %%%%%%
  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1 \\
  & \qquad \mathsf{bool}_\sim : \inUU\;\textsf{bool}\;\textsf{bool}\;
  (\lambda x_0\;x_1\; .\; x_0 \stackrel{?}{=}_{\mathbbm{2}} x_1) \\
  & \qquad \pi_\sim :
  \forall\{A_0\;A_1\;A_{0\sim}\;A_{1\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
  \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
  & \qquad\quad\quad\ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
  \{b_1 : (x_1 : A_1) \to \inU\;(B_1\;x_1)\}\\
  & \qquad\quad\quad\ \; \{a_{0\sim} : \inUU\;a_0\;a_0\;A_{0\sim}\}
  \{a_{1\sim} : \inUU\;a_1\;a_1\;A_{1\sim}\} \\
  & \qquad\quad\quad\ \; \{ b_{0\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{0\sim}\;x_0\;x_1)
    \to \inUU\;(b_0\;x_0)\;(b_0\;x_1)\;(B_{0\sim}\;x_{01}) \}\\
  & \qquad\quad\quad\ \; \{ b_{1\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{1\sim}\;x_0\;x_1)
    \to \inUU\;(b_1\;x_0)\;(b_1\;x_1)\;(B_{1\sim}\;x_{01}) \}\\
  & \qquad\quad \to \inUU\;a_0\;a_1\;A_{01\sim} \\
    & \qquad\quad \to (\forall\{x_0\;x_1\}(x_{01} : A_{01\sim}\;x_0\;x_1)
      \to \inUU\;(b_0\;x_0)\;(b_1\;x_1)\;(B_{01\sim}\;x_{01})) \\
  & \qquad\quad \to \inUU\;(\pi\;a_0\;a_{0\sim}\;b_0\;b_{0\sim})\;
      (\pi\;a_1\;a_{1\sim}\;b_1\;b_{1\sim}) \\
  & \qquad\qquad\qquad\quad\,(\lambda f_0\;f_1\;.\;\forall(x_0\;x_1)\to A_{01\sim}\;x_0\;x_1
        \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
\end{align*}}

Just as the role of $\inU$ is, as before, to classify all types that are image
of $\El$, in the same way $\inUU\ a\ a'$ classifies all relations of type $A \to
A' \to \mProp$ that are image of $\_\vdash\_\sim_{\El}\_$, given proofs $a :
\inU\ A, a' : \inU\ A'$.
%
In particular, this definition of $\inUU$ states that the appropriate
equivalence for boolean elements is the obvious syntactic equality
$\_\stackrel{?}{=}_{\mathbbm{2}}\_$, whereas functions are to be compared
pointwise.
%
Note that $\inU$ appears in the sort of $\inUU$. Since these types are
mutually defined, they form an instance of \emph{induction-induction}, a schema
that allows the definition of a type mutually with other types that contain the
first one in their signature \cite{ii}.\footnote{The main example of
  induction-induction is the intrinsic definition of a dependent type theory in
  type theory \cite{tt-in-tt}.}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  & \_\vdash\_\sim_{\El}\_ : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash x \sim_{\El} y :\equiv \ ?
\end{align*}

In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
\mProp)\ (...) \parallel$. If the type was not propositionally truncated, we could define
$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
\pi_1\,p\,x\,y$. This would make the definition of $\_\sim_\setoidU\_$ and
$\_\vdash\_\sim_{\El}\_$ in line with how we defined $\setoidU$ and $\El$.

However, this does not work in our case, since the type of $p$ \emph{is}
propositionally truncated, hence it cannot be eliminated to construct a
proof-relevant object.
%
Fortunately, we can work around this limitation by defining $p \vdash x
\sim_{\El} y$ by induction on the codes $A\ B : \setoidU$, in a way that ends
up being logically equivalent to the proposition we would have obtained by
$\pi_1\ p\ x\ y$ if there were no truncation.
%
More precisely, we need to construct proofs that for any concrete $R$ and
$\mathsf{inR}$, the types $| (R, \mathsf{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
are logically equivalent. These in turn need to be defined mutually with
$\_\vdash\_\sim_{\El}\_$. We direct the interested reader to the Agda
formalization for the full details of these definitions, as they are quite
involved.

@@

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\_,\_$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{alignat*}

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{alignat*}{10}
  & \bullet_0 && : \textsf{Con}_0 \\
  & \_,_0\_ && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{alignat*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
  & \_,_1\_ && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}
\\

>>>>

% for placeholders
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}

@@

We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
$\Sigma$-types are defined negatively by pairing $\blank,\blank$ and projections $\pi_1, \pi_2$.
We have definitional $\eta$-rules for $\Pi$-, $\Sigma$-, $\mathbf{1}$-types.
We also require indexed W-types, both in $\mType$ and $\mProp$:
$W_\Box : (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
%
From $W_{\mProp}$ we can define propositional truncation $\|\blank\| : \mType \to \mProp$, with constructor
  $|\blank| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\blank\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \blank\sim_\setoidU\blank : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \blank\vdash\blank\sim_\El\blank : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We will follow the method outlined in Section~\ref{ir-universes}. In
addition to $\inU$ for defining $\U$, we also introduce a family
$\inUU$ of binary relations between types in the universe, from which we
then define $\blank\sim_{\setoidU}\blank$.
{\small\vspace{-0.2em}
\begin{align*}
  & \textsf{data}\ \inU : \mType \to \mType_1 \\
  & \qquad \textsf{bool} : \inU\ \mathbbm{2} \\
  & \qquad \pi : \forall\{A\;A_\sim\;B\;B_\sim\}(a : \inU\;A)
  (b : (x : A) \to \inU\;(B\;x))\\
  & \qquad\ \to \inUU\;a\;a\;A_\sim \to (\forall\{x_0\;x_1\}(x_{01} : A_\sim\;x_0\;x_1)
  \to \inUU\;(b\;x_0)\;(b\;x_1)\;(B_\sim\;x_{01})) \\
  & \qquad\ \to \inU\;
  (\Sigma\;(f : (x : A) \to B\;x) \\
  &\qquad\qquad\qquad\quad\, ((x_0\;x_1 : A)(x_{01} : A_\sim\;x_0\;x_1) \to B_\sim\;x_{01}\;(f\;x_0)\;(f\;x_1))) \\
  %%%%%%
  & \textsf{data}\ \inUU : \{A\ A' : \mType\} \to \inU\ A \to \inU\ A' \to (A \to A' \to \mProp) \to \mType_1 \\
  & \qquad \mathsf{bool}_\sim : \inUU\;\textsf{bool}\;\textsf{bool}\;
  (\lambda x_0\;x_1\; .\; x_0 \stackrel{?}{=}_{\mathbbm{2}} x_1) \\
  & \qquad \pi_\sim :
  \forall\{A_0\;A_1\;A_{0\sim}\;A_{1\sim}\;A_{01\sim}\;B_0\;B_1\;B_{0\sim}\;B_{1\sim}\;B_{01\sim}\}
  \{ a_0 : \inU\;A_0 \}\{ a_1 : \inU\;A_1 \}\\
  & \qquad\quad\quad\ \; \{b_0 : (x_0 : A_0) \to \inU\;(B_0\;x_0)\}
  \{b_1 : (x_1 : A_1) \to \inU\;(B_1\;x_1)\}\\
  & \qquad\quad\quad\ \; \{a_{0\sim} : \inUU\;a_0\;a_0\;A_{0\sim}\}
  \{a_{1\sim} : \inUU\;a_1\;a_1\;A_{1\sim}\} \\
  & \qquad\quad\quad\ \; \{ b_{0\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{0\sim}\;x_0\;x_1)
    \to \inUU\;(b_0\;x_0)\;(b_0\;x_1)\;(B_{0\sim}\;x_{01}) \}\\
  & \qquad\quad\quad\ \; \{ b_{1\sim} : \forall\{x_0\;x_1\}(x_{01} : A_{1\sim}\;x_0\;x_1)
    \to \inUU\;(b_1\;x_0)\;(b_1\;x_1)\;(B_{1\sim}\;x_{01}) \}\\
  & \qquad\quad \to \inUU\;a_0\;a_1\;A_{01\sim} \\
    & \qquad\quad \to (\forall\{x_0\;x_1\}(x_{01} : A_{01\sim}\;x_0\;x_1)
      \to \inUU\;(b_0\;x_0)\;(b_1\;x_1)\;(B_{01\sim}\;x_{01})) \\
  & \qquad\quad \to \inUU\;(\pi\;a_0\;a_{0\sim}\;b_0\;b_{0\sim})\;
      (\pi\;a_1\;a_{1\sim}\;b_1\;b_{1\sim}) \\
  & \qquad\qquad\qquad\quad\,(\lambda f_0\;f_1\;.\;\forall(x_0\;x_1)\to A_{01\sim}\;x_0\;x_1
        \to B_{01\sim}\;x_{01}\;(\pi_1\;f_0\;x_0)\;(\pi_1\;f_1\;x_1))
\end{align*}}

Just as the role of $\inU$ is, as before, to classify all types that are image
of $\El$, in the same way $\inUU\ a\ a'$ classifies all relations of type $A \to
A' \to \mProp$ that are image of $\blank\vdash\blank\sim_{\El}\blank$, given proofs $a :
\inU\ A, a' : \inU\ A'$.
%
In particular, this definition of $\inUU$ states that the appropriate
equivalence for boolean elements is the obvious syntactic equality
$\blank\stackrel{?}{=}_{\mathbbm{2}}\blank$, whereas functions are to be compared
pointwise.
%
Note that $\inU$ appears in the sort of $\inUU$. Since these types are
mutually defined, they form an instance of \emph{induction-induction}, a schema
that allows the definition of a type mutually with other types that contain the
first one in their signature \cite{ii}.\footnote{The main example of
  induction-induction is the intrinsic definition of a dependent type theory in
  type theory \cite{tt-in-tt}.}

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\blank\sim_{\setoidU}\blank$ should be a $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly is not, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
\vspace{-0.2em}
\begin{align*}
  & \blank\sim_{\setoidU}\blank : \setoidU \to \setoidU \to \mProp_1 \\
  & (A , a) \sim_{\setoidU} (B , b) =
   \ \parallel \Sigma\ (R : A \to B \to \mProp)\ (\inUU\ a\ b\ R) \parallel
\end{align*}

We are now left to define the indexed equivalence relation on $\El$:
\vspace{-0.2em}
\begin{align*}
  & \blank\vdash\blank\sim_{\El}\blank : \{A\ B : \U\} \to A\ \sim_{\setoidU}\ B \to \El\ A \to \El\ B \to \mProp \\
  & p \vdash x \sim_{\El} y :\equiv \ ?
\end{align*}

In the definition above, $p$ has type $\parallel \Sigma\ (R: \El\,A \to \El\,B \to
\mProp)\ (...) \parallel$. If the type was not propositionally truncated, we could define
$p \vdash x \sim_{\El} y$ by extracting the relation out of the first component
of $p$, and apply it to $x, y$. That is, $p \vdash x \sim_{\El} y :\equiv
\pi_1\,p\,x\,y$. This would make the definition of $\blank\sim_\setoidU\blank$ and
$\blank\vdash\blank\sim_{\El}\blank$ in line with how we defined $\setoidU$ and $\El$.

However, this does not work in our case, since the type of $p$ \emph{is}
propositionally truncated, hence it cannot be eliminated to construct a
proof-relevant object.
%
Fortunately, we can work around this limitation by defining $p \vdash x
\sim_{\El} y$ by induction on the codes $A\ B : \setoidU$, in a way that ends
up being logically equivalent to the proposition we would have obtained by
$\pi_1\ p\ x\ y$ if there were no truncation.
%
More precisely, we need to construct proofs that for any concrete $R$ and
$\mathsf{inR}$, the types $| (R, \mathsf{inR}) | \vdash x \sim_{\El} y$ and $R\ x\ y$
are logically equivalent. These in turn need to be defined mutually with
$\blank\vdash\blank\sim_{\El}\blank$. We direct the interested reader to the Agda
formalization for the full details of these definitions, as they are quite
involved.

@@

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\blank,\blank$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  & \blank,\blank && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{alignat*}

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{alignat*}{10}
  & \bullet_0 && : \textsf{Con}_0 \\
  & \blank,_0\blank && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{alignat*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
  & \blank,_1\blank && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}
\\
